---
title: "HW52018"
author: "Martha Wetzel"
date: "3/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Download the dataset from the Canvas folder. Then let's load everything in:

```{r}

# Load all packages
library(phyloseq)
packageVersion("phyloseq")
library(ggplot2) 
packageVersion("ggplot2")
library(RColorBrewer)
packageVersion("RColorBrewer")

#Load HMP data
#You will need to download the data from Canvas
# You will need to change your path

load("HMPv35.RData")

HMPv35


```

What is in here?

```{r}

summary(sample_data(HMPv35))

```

Let's subset to get only samples from Stool or Palatine_Tonsils.

```{r}

#Subset 

sub <- get_variable(HMPv35, "HMPbodysubsite") %in% c("Stool", "Palatine_Tonsils")

sample_data(HMPv35)$sub <- factor(sub)

HMPv35sub <- prune_samples(sample_data(HMPv35)$sub == TRUE, HMPv35) #prune_samples is a function for filtering phyloseq objects

summary(sample_data(HMPv35sub))

HMPv35sub

```

####Problem 1

Subset the HMPv35 object to obtain only the samples from Tongue_dorsum. Call this new object HMPv35sub2

```{r}

# Fill in your code to subset

sub_t <- get_variable(HMPv35, "HMPbodysubsite") %in% "Tongue_dorsum" 

sample_data(HMPv35)$sub_t <- factor(sub_t)

HMPv35sub2 <- prune_samples(sample_data(HMPv35)$sub_t == TRUE, HMPv35) #prune_samples is a function for filtering phyloseq objects

summary(sample_data(HMPv35sub2))

HMPv35sub2

# Call the new object HMPv35sub2

```

Let's look at diversity

```{r}

#This changes the background to my favorite
theme_set(theme_bw())

#The following commands plot the different diversity measures for the two different body sites.
alpha_meas = c("Observed", "Chao1", "ACE", "Shannon", "Simpson", "InvSimpson")
(p <- plot_richness(HMPv35sub, "HMPbodysubsite", measures=alpha_meas))

```

Let's add a box plot to these displays of diversity

```{r}

p + geom_boxplot(data=p$data, aes(x=HMPbodysubsite, y=value, color=NULL), alpha=0.1) #aes stands for aesthetic mappings

```

####Problem 2

Produce the geometric box plot of diversity measures for your object, HMPv35sub2
```{r}

#Place your code for the box plot for diversity here

p <- plot_richness(HMPv35sub2, "HMPbodysubsite", measures=alpha_meas)
p + geom_boxplot(data=p$data, aes(y=value, color=NULL), alpha=0.1) + ggtitle("Richness at Tongue Dorsum")

```


Let's see what phyla are represented here. This will rely on Base R graphics.

```{r}

#Calclate number of taxa in object
ntaxa(HMPv35sub)

par(mar = c(10, 4, 4, 2) + 0.1) # make more room on bottom margin - par is a function that sets or queries graphical values
N <- 20
barplot(sort(taxa_sums(HMPv35sub), TRUE)[1:N]/nsamples(HMPv35sub), las=2) #The sort part of this statement and the [1:20] is to select the 20 most common

```


Since there are over 45000 taxa represented in this object, we are only going for the top 20.

####Problem 3

You see what taxa are most prevalent in your subset, HMPv35sub2

```{r}

#Place your code here
ntaxa(HMPv35sub2)

par(mar = c(10, 4, 4, 2) + 0.1) # make more room on bottom margin - par is a function that sets or queries graphical values
N <- 20

highval <- max(sort(taxa_sums(HMPv35sub2), TRUE)[1:N]/nsamples(HMPv35sub2))
highval

# This correctly sets y-axis range
barplot(sort(taxa_sums(HMPv35sub2), TRUE)[1:N]/nsamples(HMPv35sub2), las=2, main = 
          "Most Common Taxa for Tongue Dorsum", ylim=c(0,250))


# Note- taxa_sums returns the total # individuals from each species/taxa/OTU

```


Let's filter OTU's a bit, throwing out rare taxa. 

```{r}

#The next step filters out taxa with low occurrence throughout all samples

HMPv35subsub = filter_taxa(HMPv35sub, function(x) sum(x > 3) > (0.2*length(x)), TRUE)

# The next step filters out all taxa that occur in less than .01% of samples

HMPv35subr  <- transform_sample_counts(HMPv35subsub, function(x) x / sum(x) )
HMPv35subfr <- filter_taxa(HMPv35subr, function(x) mean(x) > 1e-5, TRUE)


HMPv35subfr

#Finally subset to only bacteria in the phylum Bacteroidetes

HMPv35subfrbac = subset_taxa(HMPv35subfr, Phylum=="Bacteroidetes")
HMPv35subfrbac

```

So we have gone from 45336 taxa in 4743 in the original HMPv35 dataset

to 45336 taxa in 631 samples in the Stool and Palatine Tonsil samples subset

to 131 taxa in 631 samples after the two filtering steps 

to 62 taxa in the phylum Bacteroidetes in 631 samples.

####Problem 4

Using your HMPv35sub2 object, throw the rare taxa out of that object, then reduce to only taxa in the phylum Bacteroidetes. Call this new object HMPv35sub2frbac 

```{r}

#Insert your code to throw out the rare taxa from the HMPv35sub2 object
#The next step filters out taxa with low occurrence throughout all samples

HMPv35sub2sub = filter_taxa(HMPv35sub2, function(x) sum(x > 3) > (0.2*length(x)), TRUE)

# The next step filters out all taxa that occur in less than .01% of samples

HMPv35subr2  <- transform_sample_counts(HMPv35sub2sub, function(x) x / sum(x) )
HMPv35subfr2 <- filter_taxa(HMPv35subr2, function(x) mean(x) > 1e-5, TRUE)


HMPv35subfr2

#Finally subset to only bacteria in the phylum Bacteroidetes

HMPv35sub2frbac = subset_taxa(HMPv35subfr2, Phylum=="Bacteroidetes")
HMPv35sub2frbac

#Call the object HMPv35sub2frbac

```

Let's look at the top taxa again in this filtered object.

```{r}

#Calclate number of taxa in object
ntaxa(HMPv35subfrbac)

par(mar = c(10, 4, 4, 2) + 0.1) # make more room on bottom margin
N <- 20
barplot(sort(taxa_sums(HMPv35subfrbac), TRUE)[1:N]/nsamples(HMPv35subfrbac), las=2)

```

####Problem 5
Using your HMPv35sub2frbac object, what is the distribution of the top 20 OTU's?

```{r}

#Place your code here.
ntaxa(HMPv35sub2frbac)

par(mar = c(10, 4, 4, 2) + 0.1) # this has to be done each time you make a graph or it resets
N <- 20
barplot(sort(taxa_sums(HMPv35sub2frbac), TRUE)[1:N]/nsamples(HMPv35sub2frbac), las=2, main = "Top 20 OTUs for Tongue Dorsum", ylim=c(0,0.03))

```


Now let's summarize this slice of data with some more graphics

```{r}

title = "plot_bar; by site; Bacteroidetes only"
plot_bar(HMPv35subfrbac, "HMPbodysubsite", "Abundance", title=title)

```

####Problem 6

Plot sample abundance by body site for your object HMPv35sub2frbac

```{r}

#Place your code for the plat here
title = "plot_bar; by site; Bacteroidetes only at Tongue Dorsum"
plot_bar(HMPv35sub2frbac, "HMPbodysubsite", "Abundance", title=title)

```


Let's try to see if there is something more to differentiate here beyond the black blobs.

```{r}

plot_bar(HMPv35subfrbac, "HMPbodysubsite", "Abundance", "Phylum", title=title)

```

####Problem 7

You try it with your HMPv35sub2frbac object

```{r}

#Place your code for your plot with family here
plot_bar(HMPv35sub2frbac, "HMPbodysubsite", "Abundance", "Phylum", title=title)

```


We can use the techniques of network science to illustrate how similar or distant samples are.

Using our filtered object with all phyla, we use the `plot_net` function to illustrate the "proximity" (or similarity) of samples, while denoting site and the sex of the participant.

```{r}
plot_net(HMPv35subfr, maxdist = 0.3, color = "HMPbodysubsite", shape="sex")
```

Let's redo switching which variable is colored and which variable is differentiated by shape.

```{r}
plot_net(HMPv35subfr, maxdist = 0.3, shape = "HMPbodysubsite", color="sex")
```
####Problem 8

Which method of display do you like best and why?

```{r}

# For these data, I think that color-coding the body subsite makes more sense, since it allows us to easily see the clustering by site. When the sites are differentiated by by shape instead of color, it is much harder to see the site clusters. However, I can imagine a situation in which it may be helpful to color-code by some kind of demographic characteristic, such as work place or family.

```
